### 2022-09-24 chapters/optimization/python-dir/python.tex
update : python chapter

\section{Introduction}
% TODO : MOVE this to SOA

% new structure : 
%  introduction  / motivation -> definition of python -> the plan of experiments -> experiments 
Dynamic programming languages, except Perl, have surpassed compiled programming languages in terms of popularity among software system developers over the past decade (cf. \Cref{fig:pypl}).
However, it remains unclear whether this category of dynamic programming languages can truly compete with compiled ones in terms of power consumption.

\begin{figure}[htbp]
    \includegraphics[width=\linewidth]{imgs/programminglanguangespopularity.png}
    \caption{PYPL Popularity of Programming Languages~\cite{noauthor_pypl_2018}.}
    \label{fig:pypl}
\end{figure}

@@

In this chapter, we explore the oblivious optimizations that can be applied to Python legacy applications to reduce their energy footprint.
As Python is widely adopted by software services deployed in public and private cloud infrastructures, we believe that our contributions will benefit a wide diversity of legacy systems and not only favorably contribute to reducing the carbon emissions of ICT, but also reduce their cloud invoice for the resources consumed by these services.
More specifically, this chapter focuses on runtime optimizations that can be adopted by developers to leverage the power consumption of Python applications.
We start by studying the impact of programmer choices, such as the type of data or control structures, on the global energy consumption of the execution code.
Then, we discuss other factors, such as the levels of concurrency, before we investigate other non-intrusive approaches to optimize the energy consumption of applications.
One type of optimization includes alternative interpreters and libraries that are dedicated to optimizing the code without changing its structures, such as \emph{ahead-of-time} (AOT) compilation and \emph{just-in-time} (JIT) libraries that are maintained by the community.

\section{Motivation}

@@

\subsection{Python Limits}
To reduce the energy consumption of Python, we started by targeting the main usage of this programming language, which is revealed to be data science and web development.
\Cref{fig:usecase} illustrates a study published by the JetBrain company on Python developers.\footnote{\url{https://www.jetbrains.com/lp/python-developers-survey-2020}}
57\% of the respondents reported that they use Python for data science, 51\% said they are using it for web development, and around 40\% are using it for system administration.\footnote{The options in this survey were not mutually exclusive.As a result, the total of the percentages is greater than $100\%$.}

@@

In this chapter, we investigate the energy footprint of Python in its most popular domains of adoption.
We first explore the data and control structures, aiming to reveal some fundamental guidelines, as \citeauthor{hasan_energy_2016} did in~\cite{hasan_energy_2016}.
Then, we measure the energy consumption of several Python implementations to propose a non-intrusive technique to improve energy efficiency.
Therefore, this chapter will address the following research questions:
\begin{compactenum}[\indent\bf RQ\,1:]
    \item \emph{What is the energy footprint of Python when used in data science?}
    \item \emph{Are the Python guidelines energy-efficient by construction?}
    \item \emph{Can we reduce the energy consumption of Python programs without altering the source code?}
\end{compactenum}

@@

\section{Green Web Development}\label{sec:webdev}

@@

\subsection{Life-cycle of a Request in Django}
Django is a MVT (\emph{Model-View-Template}) framework, which means that it follows the MVC (\emph{Model-View-Controller}) pattern.
Figure~\ref{fig:django-life-cycle} describes the life-cycle of a request in this framework.
Whenever a request arrives in Django, it is processed by the middleware layers, one at a time.
These middleware layers are in charge of authentication, security, and so on.
Once the request is processed by these layers, it is passed to the URL router, which extracts the URL from the request and tries to match it to the defined URLs.
After getting the matching URL, the corresponding view function is called.
This function is responsible for treating the request, gathering the data, and then generating the response that will be put inside a template to be returned as a response.
As Django adopts a MVT model, it also offers an automatic way to retrieve data from the database to the view using the \emph{Object Relational Mapping} (ORM)~\cite{o2008object}.

@@

\section{Python Insights}
This section aims to formulate some actionable insights to optimize the energy consumption of Python applications without altering the source code.
Therefore, we start by extending the work of Hasan~\emph{et~al.} and Oliveira~\emph{et~al.} to the context of Python environments
\cite{hasan_energy_2016,oliveira_recommending}.

\subsection{experiment and results}
%%%% concurency 
Another field of investigation is the type of data and control structures that might impact the energy consumption.
We iterate over a list using 3 methods.
First, the classical \texttt{for(i in range(len(n)))}.
However, as we can see here, unlike other programming languages, it requires extra operations, such as determining the length of the collection and then using the iterator range.
So, we tried the more adapted version \texttt{for(element in collection)}.
Moreover, in most programming languages. the \texttt{for} loop is translated to a \texttt{while} loop ( transformation from D type to B type -- asm -- ), therefore we wanted to compare this with a \texttt{while} version.
After determining the main ways to iterate over a \texttt{loop}, we run the collection algorithms of different data types to observe if they impact the energy consumption of the code, and we repeated it for the size of the collection.

@@

\subsection{Synthesis}
This study demonstrates that the optimal way to reduce the energy consumption of Python application is to follow the guidelines and to privilege the built-in functions.

\section{Python \& Multiprocessing}

\subsubsection{Cost of Parallelism in Python}
Additionally, \Cref{fig:python_multiprocessing} reports on the energy consumption of Python applications when introducing parallelism.
Python applications are single-threaded systems constrained by the \emph{global lock system} (GLS).
However, due to the increase of cores/threads per CPU, many Python libraries started to take advantage of this hardware feature by allowing concurrent execution of Python processes.
For example, the multiprocessing library\footnote{\url{https://docs.python.org/3/library/multiprocessing.html}} spawns sub-processes to increase the degree of concurrency of Python applications.
As one can see in \Cref{fig:python_multiprocessing}, the energy consumption is correlated with the number of threads until reaching the limit of physical cores, when concurrent processes start to compete for the CPU.

@@

\section{Python \& Machine Learning}
Machine learning is becoming an integral part of our daily lives, growing more potent and energy-hungry each year.
As machine learning can have a significant impact on climate change, it is vital to investigate mitigation techniques.
\subsection{Experimental Protocol}
\subsubsection{Measurement context}
\paragraph{Hardware settings}
Chifflot 8 from Grid 5000's Lille site was used for all of the trials.
The machine is outfitted with two Intel Xeon Gold 6126 CPUs, each having 12 physical cores, 192 GB of RAM, and two 32 GB Tesla V100 GPUs.

\paragraph{Software settings}
For the sake of reproducibility, each experiment is done within a Docker container using Jupyter lab. These tests are run on top of a minimal version of Debian-10 to increase the accuracy of the tests by eliminating any unnecessary processes.
\subsubsection{Input Workload}
\paragraph{Models}
Several models were developed, however, only two were used in the final trials because they achieved 94 percent accuracy in an acceptable length of time.
David Page's cifar10-fast~\footnote{\url{https://github.com/davidcpage/cifar10-fast}} and Woonhyuk Baek's torch skeleton~\footnote{\url{https://github.com/wbaek/torchskeleton}} are shown here.

\paragraph{Datasets}
The CIFAR-10 dataset was the major source of data for the studies.
It is made up of 60000 32x32 color images grouped into ten categories.

@@

\subsubsection{candidates}
The experiments were run with several different CPU and GPU configurations:
\begin{itemize}
    \item with and without GPU,
    \item with and without CPU hyper-threading,
    \item different number of CPU physical cores.
\end{itemize}

\subsubsection{Key Performance Metrics}
We used Pytorch 1.10.0 to train those models and pyjoules to measure the energy consumption of the GPU and CPU.
\begin{itemize}
    \item accuracy : in \%
    \item execution time : in seconds for both the duration of each epoch and the total duration to a chieve a certain accuracy
    \item total energy consumption : for the CPU and the  GPU
\end{itemize}

\subsection{Results \& Findings}
As the model's accuracy increases, so does the energy required for the next accuracy increment.

@@

\input{python_interperters.tex}
% \subsection{threads to validity \note{missing}}
\section{Summary}:

>>>>

\section{Introduction}

Dynamic programming languages, except Perl, have surpassed compiled programming languages in terms of popularity among software system developers over the past decade (cf. \Cref{fig:pypl}).
However, it remains unclear whether this category of dynamic programming languages can truly compete with compiled ones in terms of power consumption.
\begin{figure}[htbp]
    \includegraphics[width=\linewidth]{imgs/programminglanguangespopularity.png}
    \caption{PYPL Popularity of Programming Languages~\cite{noauthor_pypl_2018}.}
    \label{fig:pypl}
\end{figure}

@@

In this chapter, we aim to reduce the energy consumption of Python code.
We first start by presenting the motivation behind our work in Section~\ref{python:sec_motivation}.
Then we provide some insights on the impact of programmer choices in the second section. To do so, we will conduct a set of experiments on different Python aspects, from data structures and iteration mechanisms to the impact of training algorithms in machine learning, passing by the impact of the concurrency strategy and the use of orms in web development.
Finally, in Section~\ref{python:sec_interpreters},we investigate other non-intrusive approaches to optimizing the energy consumption of applications by comparing multiple Python implementations, which include alternative interpreters and libraries that are dedicated to optimizing the code without changing its structures, such as ahead-of-time (AOT) compilation and just-in-time (JIT) libraries that are maintained by the community.

\section{Motivation}\label{python:sec_motivation}

@@

\subsection{Python usecases}
To reduce the energy consumption of Python, we started by targeting the main usage of this programming language, which is revealed to be data science and web development.
\Cref{fig:usecase} illustrates a study published by the JetBrain company on Python developers.\footnote{\url{https://www.jetbrains.com/lp/python-developers-survey-2020}}
57\% of the respondents reported that they use Python for data science, 51\% said they are using it for web development, and around 40\% are using it for system administration.\footnote{The options in this survey were not mutually exclusive.As a result, the total of the percentages is greater than $100\%$.}

@@

\section{Energy optimization of python on the source code level}\label{python:sec_insights}
In this section , we investigate the energy footprint of Python in its most popular domains of adoption.
We first explore the data and control structures, aiming to reveal some fundamental guidelines, as \citeauthor{hasan_energy_2016} did in~\cite{hasan_energy_2016}.
Then, we measure the energy consumption of several Python implementations to propose a non-intrusive technique to improve energy efficiency.
Therefore, this chapter will address the following research questions:
\begin{compactenum}[\indent\bf RQ\,1:]
    \item \emph{What is the energy footprint of Python when used in data science?}
    \item \emph{Are the Python guidelines energy-efficient by construction?}
    \item \emph{Can we reduce the energy consumption of Python programs without altering the source code?}
\end{compactenum}

@@

\subsection{Green Web Development}\label{sec:webdev}

@@

\subsubsection{Life-cycle of a Request in Django}
Django is a MVT (\emph{Model-View-Template}) framework, which means that it follows the MVC (\emph{Model-View-Controller}) pattern.
Figure~\ref{fig:django-life-cycle} describes the life-cycle of a request in this framework.
Whenever a request arrives in Django, it is processed by the middleware layers, one at a time.
These middleware layers are in charge of authentication, security, and so on.
Once the request is processed by these layers, it is passed to the URL router, which extracts the URL from the request and tries to match it to the defined URLs.
After getting the matching URL, the corresponding view function is called.
This function is responsible for treating the request, gathering the data, and then generating the response that will be put inside a template to be returned as a response.
As Django adopts a MVT model, it also offers an automatic way to retrieve data from the database to the view using the \emph{Object Relational Mapping} (ORM)~\cite{o2008object}.

@@

\subsection{Python Insights}
This section aims to formulate some actionable insights to optimize the energy consumption of Python applications without altering the source code.
Therefore, we start by extending the work of Hasan~\emph{et~al.} and Oliveira~\emph{et~al.} to the context of Python environments
\cite{hasan_energy_2016,oliveira_recommending}.

\subsubsection{experiment and results}
%%%% concurency 
Another field of investigation is the type of data and control structures that might impact the energy consumption.
We iterate over a list using 3 methods.
First, the classical \texttt{for(i in range(len(n)))}.
However, as we can see here, unlike other programming languages, it requires extra operations, such as determining the length of the collection and then using the iterator range.
So, we tried the more adapted version \texttt{for(element in collection)}.
Moreover, in most programming languages. the \texttt{for} loop is translated to a \texttt{while} loop ( transformation from D type to B type -- asm -- ), therefore we wanted to compare this with a \texttt{while} version.
After determining the main ways to iterate over a \texttt{loop}, we run the collection algorithms of different data types to observe if they impact the energy consumption of the code, and we repeated it for the size of the collection.

@@

\subsubsection{Synthesis}
This study demonstrates that the optimal way to reduce the energy consumption of Python application is to follow the guidelines and to privilege the built-in functions.

\subsection{Python \& Multiprocessing}

% \subsubsection{Cost of Parallelism in Python}
Additionally, \Cref{fig:python_multiprocessing} reports on the energy consumption of Python applications when introducing parallelism.
Python applications are single-threaded systems constrained by the \emph{global lock system} (GLS).
However, due to the increase of cores/threads per CPU, many Python libraries started to take advantage of this hardware feature by allowing concurrent execution of Python processes.
For example, the multiprocessing library\footnote{\url{https://docs.python.org/3/library/multiprocessing.html}} spawns sub-processes to increase the degree of concurrency of Python applications.
As one can see in \Cref{fig:python_multiprocessing}, the energy consumption is correlated with the number of threads until reaching the limit of physical cores, when concurrent processes start to compete for the CPU.

@@

\subsection{Python \& Machine Learning}
Machine learning is becoming an integral part of our daily lives, growing more potent and energy-hungry each year.
As machine learning can have a significant impact on climate change, it is vital to investigate mitigation techniques.
\subsubsection{Experimental Protocol}
\paragraph{Measurement context}
\subparagraph{Hardware settings}
Chifflot 8 from Grid 5000's Lille site was used for all of the trials.
The machine is outfitted with two Intel Xeon Gold 6126 CPUs, each having 12 physical cores, 192 GB of RAM, and two 32 GB Tesla V100 GPUs.

\subparagraph{Software settings}
For the sake of reproducibility, each experiment is done within a Docker container using Jupyter lab. These tests are run on top of a minimal version of Debian-10 to increase the accuracy of the tests by eliminating any unnecessary processes.
\subsubsection{Input Workload}
\subparagraph{Models}
Several models were developed, however, only two were used in the final trials because they achieved 94 percent accuracy in an acceptable length of time.
David Page's cifar10-fast~\footnote{\url{https://github.com/davidcpage/cifar10-fast}} and Woonhyuk Baek's torch skeleton~\footnote{\url{https://github.com/wbaek/torchskeleton}} are shown here.

\subparagraph{Datasets}
The CIFAR-10 dataset was the major source of data for the studies.
It is made up of 60000 32x32 color images grouped into ten categories.

@@

\paragraph{candidates}
The experiments were run with several different CPU and GPU configurations:
\begin{itemize}
    \item with and without GPU,
    \item with and without CPU hyper-threading,
    \item different number of CPU physical cores.
\end{itemize}

\paragraph{Key Performance Metrics}
We used Pytorch 1.10.0 to train those models and pyjoules to measure the energy consumption of the GPU and CPU.
\begin{itemize}
    \item accuracy : in \%
    \item execution time : in seconds for both the duration of each epoch and the total duration to a chieve a certain accuracy
    \item total energy consumption : for the CPU and the  GPU
\end{itemize}

\subsubsection{Results \& Findings}
As the model's accuracy increases, so does the energy required for the next accuracy increment.

@@

\input{python_interperters.tex}
% \subsection{threads to validity \note{missing}}
\skippage

\section{Summary}: 
As Python is widely adopted by software services deployed in public and private cloud infrastructures, we believe that our contributions will benefit a wide diversity of legacy systems and not only favorably contribute to reducing the carbon emissions of ICT but also reduce their cloud invoice for the resources consumed by these services.

### 2022-09-24 chapters/optimization/python-dir/python_interperters.tex
update : python chapter

\section{Python Interpreters}\label{sec:pythoninterpreters}
%%% note : add this to the section of python interpreters
Due to the lack of support for most non-conventional Python interpreters, we mainly focus on micro-benchmarks.
Except for \textsc{PyPy}, most of the Python implementations do not support extra Python libraries, despite those extra implementations being developed to optimize a specific library, such as \textsf{Numba} with \textsf{Numpy}, or \textsf{intelpython} with machine learning algorithms.

\subsection{Preliminary studies}
For the first studies, we used the official version of Python, because the goal was mainly to highlight the impact of the structure of code on energy consumption.
One main drawback of the previous method is the work to be done to update the existing code base to reduce  energy consumption.
To avoid such hustle, we tried to find a non-intrusive approach to make the Python code more eco-friendly without altering its structure.
Python is an interpreted language, which led many initiatives to implement their own interpreter to improve one or many aspects of the Python code.
In the following section, we discuss the impact of those implementations on the energy consumption of python programs, and in which case, one should use a non-conventional interpreter to save the energy consumption of their application.

@@

\subsection{Results \& Findings}
This section will be dedicated to the results of the experiments and statistical analysis of these results.
Because we are conducting a comparison study, we first did a Shapiro-Wilk~\cite{shapiro1968comparative} test to see if the data follows a normal distribution.
Despite the fact that some implementations had a $p$-value higher than $a=0.05$ (such as CPython and ActivePython), other implementations like PyPy and Numba presented a $p$-values smaller than 0.01, which leads to reject the hypothesis $H0$ that all the distributions follow a normal distribution.
Therefore, to compare the different implementations, we performed the non-parametric Mann-Whitney U test~\cite{zimmerman1987comparative}, with CPython2 and CPython3 as base references.

The Mann-Whitney U test results for each implementation are shown in Table \ref{table:manwithyu}. The first column displays the implementation's average energy usage, while the second provides the p-value of the test when compared to the reference Python implementation. If the p-value is less than 0.05, the test result is significant. This implies that the given implementation's energy consumption differs significantly from the reference implementation's.As one can notice, most of the implementations differ significantly from the reference implementation.
\input{table-envs-based-cpython.tex}

As we want to study the position of these implementations with regards to the references, we proceed with agglomerative \emph{hierarchical cluster analysis} (HCA)~\cite{johnson1967hierarchical}.
HCA is a method that groups the different implementations based on their energy consumption.
This technique is a bottom-up approach as it starts with each implementation being a single cluster, and then merges the two most similar clusters until all the clusters are merged into a single cluster.
The similarity between two clusters is measured by the maximum distance between their furthest points.
The distance between two points is calculated as the Euclidean distance between the two points.
Figure~\ref{fig:dendogram} displays the Dendogram produced by the HCA.

@@

Except GraalPython, one can notice 3 main clusters.
The first cluster contains the reference implementations (CPython2 and CPython3) and the implementations that are based on interpreters (IronPython, IntelPython, ActivePython, Jython, and Nuitka).
Moreover, each implementation is the closest to its reference Python version.
The interpreters that are based on other virtual machines, such as Jython with JVM and IronPython with .NET, behave slightly differently from the reference implementation.
As for the MicroPython, It is the furthest from the references due to his behavior toward exceptions.
The second cluster groups the interpreter implementations that are based on a JIT compiler (PyPy2 and PyPy3).
Closer to them we find Shedskin, which is a translator that converts the Python code into C++ code to be later compiled into binary code.
Table~\ref{table:manwithyu} shows that this cluster is by far the most energy-efficient one.
While Numba2 and Numba3 are also JIT libraries, they differ from PyPy due to their manual optimization.
Unlike PyPy, which will decide by itself which part of Python to optimize.
In Numba, the developer must target the portion of the code that will be passed to JIT, while the rest of the code will be executed by the regular interpreter (aka CPython).

\paragraph{Remark:}
Because GraalPython was still in its early stages, some techniques required an abnormally lengthy time, affecting the clustering algorithm.

To help practitioners to choose the best implementation for their use case, we have created a chart that describes the behavior of each implementation toward a specific aspect of programming.
Figure~\ref{fig:tommi_all} introduces a radar plot for all the implementations that have been used in this experimentation.

\begin{figure}[!htb]
      \centering
      \includegraphics[width=1.0\linewidth]{imgs/alltomti_performance}
      \caption{Different interpreter's energy scores.}
      \label{fig:tommi_all}
\end{figure}

This graph summarizes each implementation energy score when executing our benchmark.
The lower the energy consumption, the higher the score.
We adopted a logarithmic scale to help the practitioners to compare numerous implementations due to the large differences between them.

As one can notice in Figure~\ref{fig:tommi_all}, there is no evolution between CPython2 and CPython3.
Moreover, IntelPython, and ActivePython all behave similarly to the reference implementation.
Therefore, one can conclude that the work done on those interpreters is primarily to improve a specific purpose and not the core interpreter.
ActivePython states that their version is focused on security and prepackaged libraries, which explains why it is slower than other versions due to the addition of this security support.\footnote{\url{https://www.activestate.com/solutions/why-activestate/}}
As for IntelPython, it is designed for machine learning.
Unfortunately, the \textsc{Tommti} benchmark is geared for general-purpose programming.
Despite the fact that Nuitka is a compiler, there was no discernible difference in energy consumption.
It was even more similar to the reference Python implementation than other interpreters.
However, if we look at the Nuitka techniques, we can see that they simply encapsulate the Python code with an interpreter into a single executable.
Finally, Shedskin reports on the best energy consumption pattern when it comes to arithmetic operations.
One can conclude it is due to the fact of the native type of the variables, unlike in the JIT, where they are treated as objects in the beginning.

@@

Unlike the previous interpeters, GraalPython exhibits a certain promise when it comes to complex algorithms (nested loops in particular).
% Finally, MicroPython is dedicated to embedded systems, so claiming it is a powerful CPU machine will be misleading.

@@

\subsection{Python \& Binary Operations}
The second part of this study is to look at how different Python implementations work when executing different binary operations.

\begin{figure}[!hbt]
      \small
      \center
      \begin{tabular}{|lrrrrr||r|}
            \toprule
            benchmark    & Addition & Right Rotation & Left Rotation & Or      & XOR     & Average     \\
            \midrule
            ActivePython & 676.980  & 763.208        & 651.783       & 743.016 & 728.828 & 712.76      \\
            CPython2     & 441.082  & 435.886        & 430.846       & 415.247 & 419.081 & 428.43      \\
            cpython3     & 595.209  & 685.085        & 563.839       & 657.972 & 655.560 & 531.53      \\
            \hline
            Cython       & 35.077   & 182.688        & 274.177       & 34.868  & 34.504  & 112.26      \\
            Nuitka       & 33.260   & 32.980         & 33.256        & 33.472  & 33.030  & 33.2        \\
            \hline
            Numba2       & 9.102    & 8.411          & 9.460         & 9.375   & 9.755   & \best{9.22} \\
            Numba3       & 9.566    & 10.144         & 9.219         & 9.344   & 9.665   & \best{9.59} \\
            PyPy2        & 8.456    & 7.844          & 8.286         & 8.138   & 7.952   & \best{8.13} \\
            PyPy3        & 7.552    & 8.093          & 8.108         & 8.669   & 8.623   & \best{8.21} \\
            Shedskin     & 8.024    & 8.070          & 8.399         & 8.126   & 8.277   & \best{8.18} \\
            \bottomrule
      \end{tabular}
      \caption{Energy consumption of different implementations using bit operations benchmark (in Joules).}
      \label{table:bitops}
\end{figure}

Table~\ref{table:bitops} provides a detailed report on the energy consumption of different implementations of the code.
Besides \emph{Cython}, each implementation treats the binary operation the same.
As one can see, Table~\ref{table:bitops} shows no difference in the energy consumption of those operations.
However, when it comes to Cython, the rotation operations tend to consume 5 times more energy than the other operations, and left rotations consume even more than the right ones.
The reason behind such behavior is the fact that Cython does not have a native implementation for these two operations, so it replaces them with more complex operations, which leads to an overhead of $6s$ for the right rotation and $10s$ for the left, compared to $1.5s$ for the other operations.
Therefore, it is safe to summarize all the operations into one that we call \emph{binary}.

\begin{figure}
      \centering
      \includegraphics[width=\linewidth]{imgs/bitopts_mean}
      \caption{Summary of the binary operations for different python implementations }
      \label{fig:bitops}
\end{figure}

Figure~\ref{fig:bitops} depicts the average energy consumption of different implementations for the binary operations.
One can notice 3 clusters.
The first, which is the interpreted one, uses approximately 500 Joules per operation.
The second is a non-optimized compilation that includes Cython and Nuitka and uses 10 times less energy than the first.
And the final one, which has been fully optimized, uses 50 percent less energy. This category includes JIT libraries, also known as Numba, Python implementations that include JIT, also known as PyPy, and Shedskin, which transpile the source code into C and then use a C/C++ compiler to generate the binary, as opposed to Cython and Nuitka, which directly compile the Python code.
To conclude this experiment, JIT has a huge impact when it comes to binary operations, and an inadequate runtime selection can lead to a 100x increase in energy consumption.
Furthermore, we observed that different binary operations exhibit the same energy footprint for all the implementations, except Cython.

\subsection{Conclusion}
One can conclude that the choice of Python interpreter has a significant impact on the programs' energy consumption.
This investigation is made more intriguing by the absence of a universal solution.
The primary downside is the incompatibility of some of these solutions, which causes us to make concessions when we need a generic answer.

>>>>

\section{Python Interpreters}\label{python:sec_interpreters}
For the first studies, we used the official version of Python, because the goal was mainly to highlight the impact of the structure of code on energy consumption.
One main drawback of the previous method is the work to be done to update the existing code base to reduce  energy consumption.
Despite the availability of numerous tools for code refactoring and automatic code optimization, the optimization process remains a manual effort. This is due to the fact that code optimization is a task that requires extensive understanding of both the codebase and the language itself.

To avoid such hustle, we tried to find a non-intrusive approach to make the Python code more eco-friendly without altering its structure.
Python is an interpreted language, which led many initiatives to implement their own interpreter to improve one or many aspects of the Python code. 
Therefore, nn this section, we will examine at how different Python interpreters affect the energy consumption of software,and in which case, one should use a non-conventional interpreter to save the energy consumption of their application.

\subsection{Python runtimes}\label{python:sec_interpreters}

@@

\subsection{Results \& Findings}
This part will be dedicated to the results of the experiments and statistical analysis of these results.
First and foremost, we want to know whether our samples have a normal distribution. As a result, we ran a Shapiro-Wilk~\cite{shapiro1968comparative} test on our data.
Despite the fact that some implementations had a $p$-value higher than $a=0.05$ (such as CPython and ActivePython), other implementations like PyPy and Numba presented a $p$-values smaller than 0.01, which leads to reject the hypothesis $H0$ that all the distributions follow a normal distribution.
Therefore, to compare the different implementations, we performed the non-parametric Mann-Whitney U test~\cite{zimmerman1987comparative}, with CPython2 and CPython3 as base references.

The Mann-Whitney U test results for each implementation are shown in Table \ref{table:manwithyu}. The first column shows the average energy usage of the implementation, while the second shows the p-value of the test when compared to the reference Python implementation. If the p-value is less than 0.05, the test result is significant, indicating that the energy consumption of the given implementation differs significantly from that of the reference implementation.
As can be seen, the majority of the implementations deviate significantly from the reference implementation.
\input{table-envs-based-cpython.tex}

As we want to study the position of these implementations with regards to the reference implementation (aka Cpython2 and Cpython3), we proceed with agglomerative \emph{hierarchical cluster analysis} (HCA)~\cite{johnson1967hierarchical}.
HCA is a method that groups the different implementations based on their energy consumption.
This technique is a bottom-up approach as it starts with each implementation being a single cluster, and then merges the two most similar clusters until all the clusters are merged into a single cluster.
The similarity between two clusters is measured by the maximum distance between their furthest points.
The distance between two points is calculated as the Euclidean distance between the two points.
Figure~\ref{fig:dendogram} displays the Dendogram produced by the HCA.

@@

Except GraalPython, one can notice 3 main clusters.
The first cluster contains the reference implementations (CPython2 and CPython3) and the implementations that are based on interpreters (IronPython, IntelPython, ActivePython, Jython, and Nuitka).
Moreover, each implementation is the closest to its reference Python version.
The interpreters that are based on other virtual machines, such as Jython with JVM and IronPython with .NET, behave slightly differently from the reference implementation.
As for the MicroPython, It is the furthest from the references due to his behavior toward exceptions.
The second cluster groups the interpreter implementations that are based on a JIT compiler (PyPy2 and PyPy3).
Closer to them we find Shedskin, which is a translator that converts the Python code into C++ code to be later compiled into binary code.
Table~\ref{table:manwithyu} shows that this cluster is by far the most energy-efficient one.
While Numba2 and Numba3 are also JIT libraries, they differ from PyPy due to their manual optimization.
Unlike PyPy, which will decide by itself which part of Python to optimize.
In Numba, the developer must target the portion of the code that will be passed to JIT, while the rest of the code will be executed by the regular interpreter (aka CPython).
\bigskip

%  Switch this to conclusion 
To help practitioners to choose the best implementation for their use case, we have created a chart that describes the behavior of each implementation toward a specific aspect of programming.
Figure~\ref{fig:tommi_all} introduces a radar plot for all the implementations that have been used in this experimentation.

\begin{figure}[!htb]
      \centering
      \includegraphics[height=\textheight,keepaspectratio]{imgs/alltomti_performance}
      \caption{Different interpreter's energy scores.}
      \label{fig:tommi_all}
\end{figure}

This graph summarizes each implementation energy score when executing our benchmark.
The lower the energy consumption, the higher the score.
We adopted a \textsc{logarithmic scale} to help the practitioners to compare numerous implementations due to the large differences between them.

As one can notice in Figure~\ref{fig:tommi_all}, there is no evolution between CPython2 and CPython3.
Moreover, IntelPython, and ActivePython all behave similarly to the reference implementation.
Therefore, one can conclude that the work done on those interpreters is primarily to improve a specific purpose and not the core interpreter.
ActivePython states that their version is focused on security and prepackaged libraries, which explains why it is slower than other versions due to the addition of this security support.\footnote{\url{https://www.activestate.com/solutions/why-activestate/}}
As for IntelPython, it is designed for machine learning.
Unfortunately, the \textsc{Tommti} benchmark is geared for general-purpose programming.
Despite the fact that Nuitxka is a compiler, there was no discernible difference in energy consumption.
It was even more similar to the reference Python implementation than other interpreters.
However, if we look at the Nuitka techniques, we can see that they simply encapsulate the Python code with an interpreter into a single executable.
Finally, Shedskin reports on the best energy consumption pattern when it comes to arithmetic operations.
One can conclude it is due to the fact of the native type of the variables, unlike in the JIT, where they are treated as objects in the beginning.

@@

Unlike the previous interpeters, GraalPython exhibits a certain promise when it comes to complex algorithms (nested loops in particular). However, it is still in its early stages of development and it is not yet ready for production. Since some benchmarks required an abnormally lengthy time, such as array manipulation and sorting.

% Finally, MicroPython is dedicated to embedded systems, so claiming it is a powerful CPU machine will be misleading.

@@

% \subsection{Python \& Binary Operations}
% The second part of this study is to look at how different Python implementations work when executing different binary operations.

% \begin{figure}[!hbt]
%       \small
%       \center
%       \begin{tabular}{|lrrrrr||r|}
%             \toprule
%             benchmark    & Addition & Right Rotation & Left Rotation & Or      & XOR     & Average     \\
%             \midrule
%             ActivePython & 676.980  & 763.208        & 651.783       & 743.016 & 728.828 & 712.76      \\
%             CPython2     & 441.082  & 435.886        & 430.846       & 415.247 & 419.081 & 428.43      \\
%             cpython3     & 595.209  & 685.085        & 563.839       & 657.972 & 655.560 & 531.53      \\
%             \hline
%             Cython       & 35.077   & 182.688        & 274.177       & 34.868  & 34.504  & 112.26      \\
%             Nuitka       & 33.260   & 32.980         & 33.256        & 33.472  & 33.030  & 33.2        \\
%             \hline
%             Numba2       & 9.102    & 8.411          & 9.460         & 9.375   & 9.755   & \best{9.22} \\
%             Numba3       & 9.566    & 10.144         & 9.219         & 9.344   & 9.665   & \best{9.59} \\
%             PyPy2        & 8.456    & 7.844          & 8.286         & 8.138   & 7.952   & \best{8.13} \\
%             PyPy3        & 7.552    & 8.093          & 8.108         & 8.669   & 8.623   & \best{8.21} \\
%             Shedskin     & 8.024    & 8.070          & 8.399         & 8.126   & 8.277   & \best{8.18} \\
%             \bottomrule
%       \end{tabular}
%       \caption{Energy consumption of different implementations using bit operations benchmark (in Joules).}
%       \label{table:bitops}
% \end{figure}

% Table~\ref{table:bitops} provides a detailed report on the energy consumption of different implementations of the code.
% Besides \emph{Cython}, each implementation treats the binary operation the same.
% As one can see, Table~\ref{table:bitops} shows no difference in the energy consumption of those operations.
% However, when it comes to Cython, the rotation operations tend to consume 5 times more energy than the other operations, and left rotations consume even more than the right ones.
% The reason behind such behavior is the fact that Cython does not have a native implementation for these two operations, so it replaces them with more complex operations, which leads to an overhead of $6s$ for the right rotation and $10s$ for the left, compared to $1.5s$ for the other operations.
% Therefore, it is safe to summarize all the operations into one that we call \emph{binary}.

% \begin{figure}
%       \centering
%       \includegraphics[width=\linewidth]{imgs/bitopts_mean}
%       \caption{Summary of the binary operations for different python implementations }
%       \label{fig:bitops}
% \end{figure}

% Figure~\ref{fig:bitops} depicts the average energy consumption of different implementations for the binary operations.
% One can notice 3 clusters.
% The first, which is the interpreted one, uses approximately 500 Joules per operation.
% The second is a non-optimized compilation that includes Cython and Nuitka and uses 10 times less energy than the first.
% And the final one, which has been fully optimized, uses 50 percent less energy. This category includes JIT libraries, also known as Numba, Python implementations that include JIT, also known as PyPy, and Shedskin, which transpile the source code into C and then use a C/C++ compiler to generate the binary, as opposed to Cython and Nuitka, which directly compile the Python code.
% To conclude this experiment, JIT has a huge impact when it comes to binary operations, and an inadequate runtime selection can lead to a 100x increase in energy consumption.
% Furthermore, we observed that different binary operations exhibit the same energy footprint for all the implementations, except Cython.

\subsubsection{threads to validity }
Due to the lack of support for most non-conventional Python interpreters, we mainly focus on micro-benchmarks.
Except for \textsc{PyPy}, most of the Python implementations do not support extra Python libraries, despite those extra implementations being developed to optimize a specific library, such as \textsf{Numba} with \textsf{Numpy}, or \textsf{intelpython} with machine learning algorithms.
Furthermore, we excluded from this study runtimes that were unable to complete our benchmarks because we judged that this was the bare minimum for an implementation to be employed in a real-world application.
Besides this, since the goal of this study was to provide a non-intrusive technique to decrease the energy consumption of software, we opted to omit some of the implementations that required extra work on the sourcecode to run on them, such as \textsf{Cython}.

\subsection{Conclusion}

One can conclude that the Python interpreter used has a considerable impact on the energy consumption of the programs. Furthermore, the absence of a universal solution makes this invistigation more intriguing.
While Python2 was employed by the majority of the candidates, Python3 provided better compatibility coverage.
Except for Pypy, a standout interpreter, the type of optimization strategy had a substantial impact on energy usage because each class was grouped together regardless of Python version.
Finally The use of JIT, a technology that converts code into machine code at runtime, was responsible for the significant gain in energy efficiency. The usage of translators such as Shedskin confirmed the effect of compiling Python code into machine code. Hence, making PyPy, an interpreter that has a built-in JIT, within the same class as the complied ones such as Shedskin
